Testing traversability of protein tunnels with flexible ligands using motion planning

### 2017-06-14 paper.tex
alg + planning sota

\institute{
    V. Von\'{a}sek, Martin Saska \at 
    Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}  
\and
B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno, \\
\and 
\inst{3}
TODO DAVID \at TODO DAVID
}

@@

The task of trajectory computation for the non-spherical ligand can be formulated as a path search 
in the related configuration space $\C$.
The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
ligand with $n$ DOFs.
%We propose to employ motion planning technique to determine possible trjaectories for a given ligand in the detected tunnels.
%This can be solved by search a high-dimensional configuration space $\C$. % as both 3D translation, rotation possibly with additional DOFs caused
%by the ligand's flexiblity need to be considered.
%The dimension of the configuration space $\C$ is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
%$n$ rotational bonds).
%The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
%The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
%to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
%tunnel, respectively.
%Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
The idea of sampling-based motion planning is to randomly sample the configuration space.
The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
A path in the roadmap then represents a motion in the workspace.
Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
related to 
protein folding~\cite{al2012motion,gipson2012computational,cortes2010simulating,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},analysis of loop motions~\cite{cortes2004geometric}, and even for tunnel detection~\cite{vonasek2016application,vonasek2017tunnel}.
%none of the existing work focuses on the trajectory generation in tunnels.

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query motion planning method.
RRT incrementally builds a configuration tree $\T$ rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

@@

In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnels.
As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
It can be assume that both the protein and the ligand adapt to each other and therefore.
To enable the planner find a solution, the radii of the ligand, and possibly also the radii of the protein atoms, are down-scaled
by a given factor.

The most related work to this paper is the MoMa-LigPath~\cite{cortes2005path}, that is based on the ML-RRT method~\cite{cortes2010simulating,cortes2007mlrrt}.
In comparison to our solution, the method~\cite{cortes2005path} aims for find any path leading towards the active site without
possibility to select a tunnel of interest.

\red{comparison to momaligpath~\cite{cortes2005path} + \cite{guieysse2008structure} }

Ligands are small molecules interacting with the protein. 
Due to the interaction, the relative positions of the ligand's atoms can change.
The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
After the torsional angles are set, 3D position of each atom of the ligand is computed.
By changing the angles, various conformations can be generated.
However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexiblity of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
We propose a RRT-based method to compute the trajectories of the ligand.
As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
suffer from the well known narrow passage problem~\cite{hannaWIS}.

To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
Instead, they are generated using the spheres of the tunnel.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

@@

The random samples $\qrand$ are generated around the virtual goal.
The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
The radius $R_v$ is set as the absolut distance from the tunnel centerline, e.g. $R_v=3\AA$.
Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.

@@

The configruation tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
one new configuration can be added.
The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
The collision detection can be evaluated fast using hiearchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due to the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories, and we compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
\red{ i.e., the conditional probability we discussed in the mail?}
\end{itemize}
During visualization, we map these properties onto a tunnel's surface so that they can be visually correlated with, e.g., the tunnel's shape.

\red{can you please tell in detail how is the accesibility/liugand/throughput mapped to the tunnel surface/spheres? }

@@

>>>>

\institute{
    V. Von\'{a}sek, Martin Saska \at 
    Faculty of Electrical Engineering,  
Czech Technical University in Prague, 
Technick\'a 2, 166 27, Prague 6, Czech Republic
\email{vonasek@labe.felk.cvut.cz}  
\and
B. Kozl\'{\i}kov\'{a}, A. Jur\v{c}\'{\i}k \at
Faculty of Informatics,  
Masaryk University, 
Botanick\'a 68a, 602 00 Brno, \\
\and 
D. Bedn\'a\v{r} \at
Loschmidt Laboratories, Department of Experimental Biology, Masaryk University, Brno, Czech Republic
    International Centre for Clinical Research, St. Anne’s University Hospital Brno, Brno, Czech Republic
}

@@

The task of trajectory computation for the non-spherical ligand can be formulated as a path search 
in the related configuration space $\C$.
The dimension of the configuration space is given my the DOF of the ligand, i.e., 6D for a rigid and 6D + $n$ for a flexible
ligand with $n$ DOFs.
%We propose to employ motion planning technique to determine possible trjaectories for a given ligand in the detected tunnels.
%This can be solved by search a high-dimensional configuration space $\C$. % as both 3D translation, rotation possibly with additional DOFs caused
%by the ligand's flexiblity need to be considered.
%The dimension of the configuration space $\C$ is given by the degrees of freedom (6D for a solid ligand and 6D+$n$ for a flexible ligand with
%$n$ rotational bonds).
%The feasible configurations, where the ligand (robot) does not touch and obstacle, form the free region $\CF$.
%The tunnel traversability can be then formulated as a path planning from the start configuration $\qinit \in \CF$ 
%to a goal configuration $\qgoal\in\CF$, where both $\qinit$ and $\qgoal$ are located at the beginning and the end of the
%tunnel, respectively.
%Sampling-based methods like PRM~\cite{kavrakiPRM} or RRT~\cite{lavalleRRT} can be used to search such a path.
The idea of sampling-based motion planning is to randomly sample the configuration space.
The random samples are classified as free or non-free using collision detection and  the free ones are stored into a roadmap.
A path in the roadmap then represents a motion in the workspace.
Beyond robotics, sampling-based motion planners have been successfully applied in computational biology in studies
related to 
protein folding~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro}, analysis of loop motions~\cite{cortes2004geometric},
 tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
 exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
%none of the existing work focuses on the trajectory generation in tunnels.

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
incrementally builds a configuration tree $\T$ rooted at the initial configuration $\qinit$.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node in the tree $\qnear \in \T$ is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

@@

A RRT-based method for computing exit pathways for a small molecule was presented in~\cite{cortes2010simulating}.
Both protein and ligand are represented as flexible kinematic chains.
To decrease the dimension of the related configuration space, the RRT-ML variant~\cite{cortes2007mlrrt} is employed.
In RRT-ML expands the tree primarily using those DOF that are essential for achieving motion of the ligand (i.e., rotation
        and translation) and it employs the other DOFs (i.e., that are responsible for conformation change) if they hinder
the growth of the tree.
The RRT-ML is also used in the MoMa-LigPath online service.

The~\cite{cortes2010simulating} approach aims to find any exit pathway for a given ligand, which requires
to search the whole configuration space of the ligand/protein complex.
This might be a possible disadvantage, as the proteins are dense structures with relatively narrow void space, which has the
properties of the narrow passage~\cite{hannaWIS}.
Although there are usually multiple pathways from a given active site, not all of them are traversable by a given ligand.
Many iterations are needed to find a solution, which increases the computational time.

The protein tunnels computed for spherical probes contain valuable information about the protein void space.
Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly
used by a ligand to exit the protein.
The biochemically relevant tunnels can be then analyzed separately.

In this paper, we propose to employ RRT to compute the trajectory for a given ligand in a given tunnel.
We assume that the tunnel is computed using the standard tunnel detection tool and is represented using a sequence of spheres.
To guide the configuration tree through the configuration space, the random samples are generated gradually using the spheres of the tunnel.
As the tunnels are usually computed on protein models without ligands, they are usually very narrow and a ligand can barely traverse them.
It can be assume that both the protein and the ligand adapt to each other and therefore.
To enable the planner find a solution, the radii of the ligand, and possibly also the radii of the protein atoms, are down-scaled
by a given factor.

This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
protein/ligand complex, but only its subspace need to be searched.
Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
again for each ligand.

%Ligands are small molecules interacting with the protein. 
%Due to the interaction, the relative positions of the ligand's atoms can change.
The ligands can be modeled as a multi-link kinematic chain where the torsional angles phi/psi can change in a given range~\cite{songPFpath}.
After the torsional angles are set, 3D position of each atom of the ligand is computed.
By changing the angles, various conformations can be generated.
However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.

@@

Proteins and ligands are represented by the hard sphere model, where the radius of each sphere (atom) is given by its van der Waals radius.
%Let $\SS \subset \mathbf{R}^3$ denote the union of all spheres, i.e., the geometry of the protein.
%We further assume that a subset of atoms located at the surface, are known. These atoms, reffered to as {\sl surface atoms} in this paper,
%   can be detected as atoms located at $\alpha$-shape of the molecule.
%The tunnels need to be found for a spherical probe $\Sprobe$ of radius $\probe$.
The flexibility of a ligand is modeled using the set $\L$ of its conformations.
The conformations can be prepared e.g., considering the potential energy using tools like Rosetta.
We assume that a conformation $l \in \L$ can be switched to all other conformations in $\L$.

@@

The task of the traversability study is to determine if a given ligand can traverse  a given tunnel in the protein.
The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
The main loop of the method is described in Alg.~\ref{alg::main}.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

@@

The random samples $\qrand$ are generated around the virtual goal.
The translation of $\qrand$ is generated from Normal distribution centered at $t_v$ with radius $R_v$.
The radius $R_v$ is set as the absolute distance from the tunnel centerline, e.g. $R_v=3\AA$.
Alternatively, $R_v$ can be set to $R_v=r_v$ according the radius of the tunnel at the virtual goal.

@@

The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
one new configuration can be added.
The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}

@@

\red{Adam: }
As described above, our approach enables to estimate the probability that a ligand given a minimum scale passes through a tunnel.
On the other hand, the probability does not describe any details about the ligand's passage, e.g., how hard it is to pass some part of the tunnel, which trajectory is the most used, etc.
Moreover, it is not feasible to directly visualize and explore the resulting trajectories due the visual clutter (see Fig.~\ref{fig:trajectories} left) caused by their high number and close proximity.
Therefore, we propose to extract and visualize the complex information about the ligand's passage through a tunnel.
In this manner, we aim to further support the decision process of a chemist about a tunnel.
In order to do that, we employ an ensemble of resulting trajectories in order to compute and visualize the following tunnel properties:
\begin{itemize}
  \item \textbf{Accessibility} --- The probability of the ligand that it reaches a tunnel's part.
  \red{ i.e., number of trajectories ending in the tunnel part over all trajectories? }
  \item \textbf{Ligand Scale} --- The scale of the ligand wheh it passes a tunnel's part.
  \item \textbf{Throughput} --- The probability of the ligand that it passes through a tunnel's part.
\red{ i.e., the conditional probability we discussed in the mail?}
\end{itemize}

\red{can you please tell in detail how is the accesibility/ligand/throughput mapped to the tunnel surface/spheres? }

@@

\section{Discussion}

guiding, parametes, different sampling, flexibility
